---
title: "IHM Differential Power on Heart Failure"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: True
  html_notebook:
    toc: yes
editor_options:
  chunk_output_type: console
---

Use the cohorts from Egerstedt et al., Nature Comm, 2019 to assess 

1) whether subjects with different cardiovascular status have different IHM

Note that demographics data are not available; assuming the group membership for the MDCS data

Extended Data Figure 7h

```{r setup}
library(tidyverse)
library(lme4)
source(file.path("scripts","utils.R"))

# IHM protein surrogate
ihm.sig <- read.csv(file.path("data","IHM","IHM.protein.signature.txt"),sep = "\t",strip.white = T,stringsAsFactors = F)
ihm.sig.pos <- gsub("[-, ]","",subset(ihm.sig,direction == "positive")$feature)
ihm.sig.neg <- gsub("[-, ]","",subset(ihm.sig,direction == "negative")$feature)

# HF cohort
mdcs.proteins <- read.csv(file.path("data","HF","Egerstedt.et.al.2019.somalogic.mdcs.cohort.csv"),strip.white = T,stringsAsFactors = F,skip = 1)
manifest.hf.proteins <- read.csv(file.path("data","HF","Egerstedt.et.al.2019.somalogic.manifest.HF.csv"),strip.white = T,stringsAsFactors = F,skip = 1)
# Four proteins with different names but are equivalent
colnames(manifest.hf.proteins) <- colnames(mdcs.proteins)
transplant.proteins <- read.csv(file.path("data","HF","Egerstedt.et.al.2019.somalogic.transplant.csv"),strip.white = T,stringsAsFactors = F,skip = 1)
pre.transplant.proteins <- subset(transplant.proteins,Timing == "PreHTX")[,-c(1:2)]
post.transplant.proteins <- subset(transplant.proteins,Timing == "PostHTX")[,-c(1:2)]
colnames(pre.transplant.proteins) <- colnames(mdcs.proteins)
colnames(post.transplant.proteins) <- colnames(mdcs.proteins)

merged.proteins <- log2(rbind(mdcs.proteins,manifest.hf.proteins,pre.transplant.proteins,post.transplant.proteins))
group=c(rep("1 -\nNested Population",583),rep("2 -\nIncident HF",185),rep("3 -\nManifest HF",85),
        rep("4 -\nPre Transplant",30),rep("5 -\nPost Transplant",30))
```

## IHM scores of subjects
```{r ihm.scores,out.width="900px"}
ihm.combined.scores <- calculate_signature_score(t(merged.proteins),ihm.sig.pos,ihm.sig.neg)
ihm.df <- data.frame(subject=group,ihm.score=ihm.combined.scores)

# only first 3 groups
ggplot(subset(ihm.df,subject %in% unique(ihm.df$subject)[1:3]),aes(subject,ihm.score)) + 
  geom_boxplot(outlier.shape = NA) + geom_jitter(height=0,width=0.2,alpha=0.6,aes(color=subject)) +
  ggpubr::stat_compare_means(comparisons = list(c(1,2),c(2,3),c(1,3)),size=3) +
  theme_bw() + ggthemes::scale_color_colorblind() + ylim(NA,1.25) +
  ggtitle("IHM using both positive and negative proteins") + ylab("IHM protein surrogate score")
#ggsave("HF.3groups.ihm.boxplot.pdf",width=5,height=3)
```

## Reproducing Fig 2. in source paper
Confirming the group membership in the MDCS data
```{r pca}
proteins.of.interest <- c("CRP","PRKACA","ProteinC","TSP2","NterminalproBNP","BLC")
merged.proteins.df <- data.frame(subject=group,merged.proteins)

for (i in proteins.of.interest) {
  print(ggplot(merged.proteins.df,aes_string("subject",paste0("2^",i))) + geom_boxplot() + theme_bw())
}
```


### Session Info
```{r session}
sessionInfo()
```

