---
title: "Figure 1"
author: "Dylan Hirsch"
date: "3/29/2019"
output: html_document
---

Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = normalizePath("../../../"))
```

Load library
```{r}
library(ggplot2)
library(ggrepel)
library(dplyr)
library(reshape2)
library(Biobase)
```

Set input paths
```{r}
if(!exists("snakemake")){
  setwd("../../..")
  source("scripts/util/paper/parse_snakemake.R")
  parse_snakemake(rule = "figure_1_statistics")
}

## Monogenic data base
METADATA.IN.PATH = snakemake@input[[1]] ## "Metadata/monogenic.de-identified.metadata.RData"
```

## KS Statistic for difference in age distributions between cases and controls in the discovery cohort

Load metadata
```{r}
load(METADATA.IN.PATH)
```

Get ages for patients in training set
```{r}
df = monogenic.all.assays %>%
  select(visit_id, patient_id, patient_age_at_time_of_blood_draw, condition, analysis_group) %>%
  mutate(patient_id = paste0('P', patient_id)) %>%
  mutate(age = patient_age_at_time_of_blood_draw) %>%
  mutate(group = ifelse(condition == "Healthy", "Control", "Case")) %>%
  filter(analysis_group == 'Discovery') %>%
  select(-patient_age_at_time_of_blood_draw, -analysis_group)
```

Average ages over various visits, first by averaging ages from samples within a visit (should all be almost exactly the same), and then averaging ages from samples across visits
```{r}
df = df %>%
  group_by(visit_id) %>%
  summarise(age = mean(age), patient_id = unique(patient_id), condition = unique(condition), group = unique(group)) %>%
  ungroup() %>%
  group_by(patient_id) %>%
  summarise(age = mean(age), condition = unique(condition), group = unique(group)) %>%
  ungroup()

```

Perform the KS test
```{r}
ks.test(df$age[df$group == 'Control'], df$age[df$group == 'Case'])
```

## Fisher's exact test for gender differences between cases and controls in the discovery cohort

Load metadata
```{r}
#load(METADATA.IN.PATH)
```

Get metadata for patients in training set
```{r}
df = monogenic.all.assays %>%
  select(patient_id, gender, condition, analysis_group) %>%
  unique() %>%
  mutate(patient_id = paste0('P', patient_id)) %>%
  filter(analysis_group == 'Discovery') %>%
  mutate(group = ifelse(condition == 'Healthy', 'Case', 'Control')) %>%
  group_by(gender, group) %>%
  summarise(total = length(patient_id)) %>%
  ungroup()
```

Convert to matrix
```{r}
X = df %>%
  dcast(gender ~ group, value.var = 'total') %>%
  select(-gender) %>%
  as.matrix()
```

Perform the fisher's test
```{r}
fisher.test(X)
```

```{r}

#dat <- monogenic.all.assays %>%
#        select(patient_id, visit_id, condition, analysis_group, assay_type) %>%
#        filter(analysis_group == "Discovery") %>%
#        filter(assay_type %in% c("Microarray", "Somalogic", ""))
```


#Number of repeat samples in each datatype

```{r}
TBNK.IN.PATH <- snakemake@input[["tbnk"]]
SOMA.IN.PATH <- snakemake@input[["soma"]]
ARRAY.IN.PATH <- snakemake@input[["array"]]
TBNK.TEST.IN.PATH <- snakemake@input[["tbnk_test"]]
SOMA.TEST.IN.PATH <- snakemake@input[["soma_test"]]
ARRAY.TEST.IN.PATH <- snakemake@input[["array_test"]]

tbnk <- readRDS(TBNK.IN.PATH)
soma <- readRDS(SOMA.IN.PATH)
array_ <- readRDS(ARRAY.IN.PATH)
tbnk_test <- readRDS(TBNK.TEST.IN.PATH)
soma_test <- readRDS(SOMA.TEST.IN.PATH)
array_test <- readRDS(ARRAY.TEST.IN.PATH)


data_list <- list(tbnk = tbnk, soma = soma, array_ = array_, tbnk_test = tbnk_test,
                   soma_test = soma_test, array_test = array_test)
```

```{r}
conditions = monogenic.all.assays[monogenic.all.assays$analysis_group %in% c('Discovery'), 'condition'] %>% unique

## Remove any unknown samples from these conditions
unknowns = c('Unknown', 'U_PI3K', 'U_Kastner', 'U_STAT1', 'U_CTLA4', 'U_Telomere')
conditions = setdiff(conditions, unknowns)

out_list <- list()
for(nm in names(data_list)){
  pdat <- pData(data_list[[nm]])
  pdat <- pdat %>% filter(condition %in% conditions)

  n_subjects <- length(unique(pdat$patient_id))
  n_visits <- length(unique(pdat$visit_id))
  n_repeats <- n_visits - n_subjects 

  n_subj_w_repeats <- sum(table(pdat$patient_id) > 1)

  out_list[[nm]] <- c(n_subjects = n_subjects, n_visits = n_visits, n_repeats = n_repeats, n_subj_w_repeats= n_subj_w_repeats)
  #print(nm)
  #print(vec)
}
mat <- do.call(rbind, out_list)
print(mat)

print("totals")
print(colSums(mat))

print("total number of unique disease patients")
pat_list <- lapply(data_list, function(x){
 pData(x) %>%
         filter(condition %in% conditions) %>%
         filter(condition != "Healthy") %>%
         pull(patient_id)

                   })
length(unique(unlist(pat_list)))

print("There were some additional patients with tbnk data, but no array/somalogic")
pats_with_tbnk_but_no_array_soma <- setdiff(unique(unlist(pat_list[c(1, 4)])), unique(unlist(pat_list[c(2,3,5,6)])))
pats_with_tbnk_but_no_array_soma

pats_with_tbnk_but_no_array_soma <- setdiff(unique(unlist(pat_list[c(1)])), unique(unlist(pat_list[c(2,3,5,6)])))
pats_with_tbnk_but_no_array_soma

print("total number of unique healthy")
healthy_list <- lapply(data_list, function(x){
 pData(x) %>%
         filter(condition == "Healthy") %>%
         pull(patient_id)

                   })
length(unique(unlist(healthy_list)))
setdiff(unique(unlist(healthy_list[c(1, 4)])), unique(unlist(healthy_list[c(2,3,5,6)])))

print("total number of unique visits")
visit_list <- lapply(data_list, function(x){
 pData(x) %>%
         filter(condition %in% conditions) %>%
         pull(visit_id)

                   })
length(unique(unlist(visit_list)))


#monogenic.all.assays %>% 
#        mutate(pid = paste0("P", patient_id)) %>%
#        filter( pid %in% pats_with_tbnk_but_no_array_soma) %>%
#        select(patient_id, assay_type, visit_id)

```

```{r}
#conditions = monogenic.all.assays[monogenic.all.assays$analysis_group %in% c('Discovery'), 'condition'] %>% unique
#
### Remove any unknown samples from these conditions
#unknowns = c('Unknown', 'U_PI3K', 'U_Kastner', 'U_STAT1', 'U_CTLA4', 'U_Telomere')
#conditions = setdiff(conditions, unknowns)
#
#counts_both_test_train = monogenic.all.assays %>%
        #filter(assay_type %in% c("Microarray", "Somalogic")| patient_id %in% c(42, 59, 78 171)) %>%
#  select(patient_id, visit_id, analysis_group, condition) %>%
#  filter(!grepl("CTRL", condition, ignore.case = T)) %>%
#  filter(!grepl("control", condition, ignore.case = T)) %>%
#  filter(condition %in% c("Healthy", conditions)) %>%
#  unique() %>% # Ensure there are no repeated visits
#  group_by(analysis_group) %>%
#  summarise(primary.samples = length(unique(patient_id)), # Count the number of primary samples for each condition
#            all.samples = length(unique(visit_id))) %>% # Count the number of total samples for each condition
#  mutate(repeat.samples = all.samples - primary.samples) %>% # Get the number of repeat samples for each condtion
#  ungroup() 
#
#counts_both_test_train
#
#colSums(as.matrix(counts_both_test_train[, -1]))
#```
#
#
#```{r}
#all_pats <- unique(c(tbnk$patient_id, array_$patient_id, soma$patient_id))
#total_pats <- length(all_pats)
#print("Total unique patients")
#print(total_pats)
```

```{r}
dat <- monogenic.all.assays %>%
        filter(analysis_group == "Discovery") %>%
        filter(assay_type %in% c("Microarray", "Somalogic"))

tab <- table(monogenic.all.assays$patient_id, monogenic.all.assays$visit_id)
dup_pats <- rownames(tab)[rowSums(tab > 0) > 1]

ranges <- dat %>% 
        filter(patient_id %in% dup_pats) %>%
        group_by(patient_id) %>%
        mutate(blood_draw_date = as.numeric(blood_draw_date)) %>%
        summarise(min_date = min(blood_draw_date), max_date = max(blood_draw_date)) %>%
        mutate(day_range = max_date - min_date) %>%
        filter(day_range > 0)

quantile(ranges$day_range)

```


