---
title: "Figure 4 statistics"
author: "Dylan Hirsch"
date: "8/9/2019"
output: html_document
---

Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = normalizePath("../../.."))
```

Load libraries
```{r}
library(Biobase)
library(dplyr)
```

Source utilites
```{r}
source('scripts/util/Groups/groups.R')
```

Set paths
```{r}
# Sample meta data for samples run in the random forest
META.IN.PATH = snakemake@input[[1]] #"Classification/healthy_random_forest_sample_meta_data_all.RDS"
# The healthy index from the healthy random forest
INDEX.IN.PATH = snakemake@input[[2]] #"Classification/results/healthy_rf_results_all.RDS"
```

# Tests for differences of HI among groups

Load data
```{r}
meta = readRDS(META.IN.PATH)
healthy.index = readRDS(INDEX.IN.PATH)
```

Load groups
```{r}
AI = util.get_ai()
PID = util.get_pid()
Telo = util.get_tert_terc()
```

Initialize data frame with only patients' healthy indexes
```{r}
df = data.frame(healthy.index = healthy.index$all.modules.plus.grey.with.tbnks, condition = meta$condition, 
                stringsAsFactors = FALSE)
df = df %>%
  filter(condition %in% c(AI, PID, Telo)) %>%
  mutate(group = condition %>%
           replace(.,. %in% AI, 'AI') %>%
           replace(.,. %in% PID, 'PID') %>%
           replace(.,. %in% Telo, 'Telo')) %>%
  mutate(condition = factor(condition, unique(condition))) %>%
  mutate(group = factor(group, c('AI','PID','Telo')))
```

## See if HI is different across conditions (not including Healthy)
```{r}
kruskal.test(df$healthy.index, df$condition)
```

## See if HI is different across groups (AI, PID, Telo)
```{r}
kruskal.test(df$healthy.index, df$group)
```

## See if HI is different across PID conditions
```{r}
df.subset = df %>% filter(group == 'PID')
kruskal.test(df.subset$healthy.index, df.subset$condition)
```

## See if HI is different across AI conditions
```{r}
df.subset = df %>% filter(group == 'AI')
kruskal.test(df.subset$healthy.index, df.subset$condition)
```

## See if HI is different across Telo conditions
```{r}
df.subset = df %>% filter(group == 'Telo')
kruskal.test(df.subset$healthy.index, df.subset$condition)
```

# Test if age distributions are similar among health and disease subjects tested

Load data
```{r}
meta = readRDS(META.IN.PATH)
healthy.index = readRDS(INDEX.IN.PATH)
```

Test for differences between healthy controls and patient distributions using a KS test
```{r}
healthy.ages = meta$Age[meta$condition == "Healthy"]
case.ages = meta$Age[meta$condition != "Healthy"]
ks.test(healthy.ages, case.ages)
```

Also do a Kruskall Wallace test for differences in ages among groups
```{r}
ages = c(healthy.ages, case.ages)
groups = factor(c(rep('healthy', length(healthy.ages)), rep('case', length(case.ages))), c('healthy','case'))
kruskal.test(c(healthy.ages, case.ages), groups)
```

# Get Variances of the healthy index for each group

Load data
```{r}
index = readRDS(INDEX.IN.PATH)
meta = readRDS(META.IN.PATH)
```

```{r}
index = index$all.modules.plus.grey.with.tbnks
sapply(unique(meta$condition), function(condition) {
  var(index[meta$condition == condition])
})
```

# Test if there is a difference in the Healthy Index of older versus younger subjects among healthy controls

Load data
```{r}
index = readRDS(INDEX.IN.PATH)
meta = readRDS(META.IN.PATH)
```

Get healthy indexes for the younger and older halves of patients
```{r}
index = index$all.modules.plus.grey.with.tbnks

median.age = median(meta$Age[meta$condition == 'Healthy'])
younger.indexes = index[meta$Age <= median.age & meta$condition == 'Healthy']
older.indexes = index[meta$Age > median.age & meta$condition == 'Healthy']
wilcox.test(younger.indexes, older.indexes)
```

# Do Wilcoxon tests of the healthy index for each group versus healthy

Load data
```{r}
index = readRDS(INDEX.IN.PATH)
meta = readRDS(META.IN.PATH)
```

```{r}
index = healthy.index$all.modules.plus.grey.with.tbnks
```

Do the wilcoxon for each condition versus healthy and display results
```{r}
conditions = setdiff(unique(meta$condition), 'Healthy')
for(condition in conditions) {
  print(condition)
  print(paste0('n = ',sum(meta$condition == condition)))
  res = wilcox.test(index[meta$condition == 'Healthy'], 
              index[meta$condition == condition], 
              alternative = 'greater')
  print(res)
}
```

