---
title: "Figure 5"
author: "Dylan Hirsch"
date: "5/2/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(root.dir = normalizePath("../../../"))
```

```{r}
library(ggplot2)
library(Biobase)
library(ggpubr)
library(MetaIntegrator)
library(limma)
```

Source utilities
```{r}
source('scripts/util/Enrichment/hyperGeo.R')
```

Load data
```{r}
# Baltimore Aging Study eset
ESET.IN.PATH = snakemake@input[[1]] #'Reference/ferrucci/processed/aging_eset.RDS'
# HI proteomic surrogate signature
SIGNATURE.IN.PATH = snakemake@input[[2]] #'Classification/proteomic_surrogates/healthy.index.plasma.surrogate.somaId.RDS'
# Baltimore Aging Study somamer associations from their paper
TABLE.IN.PATH = snakemake@input[[3]] #'Reference/ferrucci/raw/acel12799-sup-0004-TableS3.txt'
```

## Wilcoxon test to see if proteins in the table that belong to our HI signature tend to have lower pvalues than those that do not

Load data
```{r}
eset = readRDS(ESET.IN.PATH) ## The baltimore aging cohort eset
feature.data = fData(eset) ## The baltimore aging cohort eset
signature = readRDS(SIGNATURE.IN.PATH) ## The plasma somalogic surrogate signature for the HI
table = read.table(TABLE.IN.PATH, header = TRUE, sep = '\t') ## The Ferrucci data significance table
```

See if proteins in the table that belong to our HI signature tend to have lower pvalues than those that do not
```{r}
signature = unname(unlist(signature)) ## Get the somalogic proteins used in the proteomic surrorgate signature of the HI
signature = signature[signature %in% feature.data$SomaId] ## subset to only the proteins measured by Ferrucci
pvals = table$p ## Get the p values from the Ferrucci data set
names(pvals) = table$SomaId
## Test whether the pvalues of proteins in the HI surrogate signature are more significant than those outside the
## HI surrogate signature
wilcox.test(pvals[names(pvals) %in% signature], pvals[! names(pvals) %in% signature], alternative = 'less')
```

## Fisher's exact test for the overlap of significant proteins from the original Baltimore Aging Cohort analysis and our HI signature

See what the overlap in proteins is between the Ferrucci data (with cutoff) and ours
```{r}
pvals = pvals[p.adjust(pvals, 'fdr') < .05]
```

```{r}
hyperGeoTest(names(pvals), feature.data$SomaId, signature)
```

